Quantifying Limits to Detection of Early Warning for Critical Transitions

### 2011-10-14 boettiger.tex
adding figures to new appendix

The ROC curves for these data (Fig. 3) show that the summary-statistic based indicators 
frequently lack the sensitivity to distinguish reliably between observed patterns from a stable or unstable system.  
The large correlations observed in the empirical examples (Fig. 2) are not uncommon in stable systems.  
The best-performing summary statistic would, at a 5\% false positive rate, have a 10\% chance of catching a true positive 
in both the Daphnia and Glaciation data set.  
Meanwhile, our likelihood-based approach performs substantially better, 
having respectively a 30\% in Daphnia and 100\% chance in Glaciation set of catching a true positive at a 5\% false positive rate. 
Because we have the entire ROC curve, we are not restricted to an evaluation at a fixed false-positive rate such as 5\%, 
but can accept more false-positives to avoid missing as many impending catastrophes.  
This is particularly valuable in the context of critical transitions, 
where data are often limited and the consequences of missed detection can be far more severe than false alarms.

@@

@@

We stress that all of the early warning signals approaches must only be applied when the underlying assumptions can be justified.  
The summary statistics such as variance or autocorrelation are no more general than our explicit models, % wow
and systems that do not correspond to these bifurcations can exhibit sudden transitions without any warning~\cite{Hastings2010}.
For instance, bifurcations resulting in sudden transitions have been observed in ecological models in which 
the variance actually deceases as a sudden shift approaches~\cite{Schreiber2003, Schreiber2008}.

Meanwhile, if our knowledge of the system gives reason to be concerned that it may experience a saddle-node or transcritical bifurcation,

>>>>

The ROC curves for these data (Fig. 3) show that the summary-statistic based indicators 
frequently lack the sensitivity to distinguish reliably between observed patterns from a stable or unstable system. 
The large correlations observed in the empirical examples (Fig. 2) are not uncommon in stable systems. 
For instance, we can read from the curve what fraction of true positives would be detected by a variance-based indicator
set to a 5\% false positive rate: 25\% in the simulated set, 5.0\% in the \emph{Daphnia} set, and 5.4\% in the Glaciation data set.
Compare this to likelihood: 61\%, 34\%, and 100\% of the replicate true positives would be correctly identified from the respective systems.
The value of the ROC curve is that we need not restrict the analysis to a pre-specified error such as 5\%, as we would in classical hypothesis testing.  
We could instead decide that we wanted no to catch no less than 90\% of the true-positives, and read off the corresponding false positives.  
For variance this corresponds respectively to a 49\%, 81\%, and 93\% false positive rate, while for likelihood a 55\%, 35\%, 0\% false positive.\footnote{
Each false-positive, true-positive coordinate on the ROC curve corresponds to a threshold of the indicator statistic, as illustrated in Fig. 1.  
Comparing this threshold with the observed value classifies the signal as detected or undetected.  See SI text.}
This flexibility is particularly valuable in the context of critical transitions, 
where data are often limited and the consequences of missed detection can be far more severe than false alarms.

@@

%%% Model adequacy
In any application of early warning signals, it is essential to address the question of model-adequacy.  
This is every bit as important for the summary-statistic based indicators such as variance and autocorrelation
as it is for the approach taken here: it has been shown that these statistics correspond to particular assumptions about 
an underlying process, and systems that violate these assumptions exhibit sudden transitions without any warning~\cite{Hastings2010}.
Elsewhere sudden transitions have been shown in ecological models where the variance deceases rather then increases as the system nears a bifurcation point~\cite{Schreiber2003, Schreiber2008}.  
Our approach formalizes the assumptions about the underlying process to match the assumptions of the other warning signals.  
All are based around a bifurcation process in which the principle eigenvalue passes through zero.

% remove skew, CV

It may appear that our approach is more limited as the models are explcitly linear with Gaussian noise processes.
This is consistent

While

@@

We stress that all of the early warning signals approaches must only be applied when the underlying assumptions can be justified.  
The summary statistics such as variance or autocorrelation are no more general than our explicit models, % wow
and 
Meanwhile, if our knowledge of the system gives reason to be concerned that it may experience a saddle-node or transcritical bifurcation,
